### 2023-04-26 chapter0/chapter0.tex
Added a manual word break point
Added citations to chapter 2 introduction paragraph 1

Unsurprisingly, disordered regions are characterized by a relative depletion of hydrophobic residues~\cite{Uversky2000}. However, there is no simple formula which accurately predicts disorder in a protein using only the hydrophobicity values of its constituent residues. Other factors, such as the number and distribution of charged residues, also impact a sequence's predisposition for disorder and its resulting conformational class~\cite{Mao2010, vanderLee2014, Das2015}. For example, sequences with high number of either positively or negatively charged residues, called polyelectrolytes, tend to form stiff rods because the like charges repel each other. However, if a sequence contains a high number of positively and negatively charged residues in roughly equal proportion, the sequence is called a strong polyampholyte, and its conformational class depends on the distribution of those charged residues. If the two classes of residues are segregated into separate blocks of like charges, they attract and form hairpins. However, if they are evenly distributed, the attractive and repulsive forces balance on average, and the sequence generally assumes expanded coil-like conformations. High numbers of polar amino acids, which are hydrophilic but not charged, are associated with semi-compact globules. Though their side chains, the portions which give each amino acid its unique identity, are hydrophilic, their interactions with water are not sufficient to overcome the tendency of the hydrophobic backbone, composed the donor and receptor sites common to all amino acids, to self-associate. Thus, like folded domains, the structures of disordered proteins are dictated by their sequences. However, because IDRs do not make stable contacts between specific residues in their chains, multiple sequences can generally correspond to a single conformational class.

The structural diversity of IDRs is directly related to their functional plasticity, and as IDRs are not confined to one conformation, they can interact with and bind to many possible partners. Often these partners are other proteins, but they can also be other macromolecules like DNA or even small molecules and ions. As a result of this adaptability, IDRs are enriched in proteins involved in cell signaling. Because cells are highly compartmentalized, they have a dizzying array of mechanisms to relay messages.\footnote{The most fundamental compartment is the cell, which roughly separates inside from outside and life from non-life. However, the cells of more complex organisms called eukaryotes contain additional subcompartments called organelles. Compartmentalization is essential in living systems because it confines and concentrates biochemical reactions that would be harmful if they occurred at the wrong place or time. However, it also introduces many complications because information in the form of physical or chemical signals cannot travel freely.} Many are mediated by interactions between proteins, which in turn create a change in the state of the system that propagates the signal. Often these state changes are chemical alterations made by one protein to another, which are termed post-translational modifications (PTMs).\footnote{A protein's amino acid sequence is encoded in a cell's DNA, and the process of reading that information to create a protein (from an intermediate molecule called RNA) is translation. Therefore, any modifications to a protein after its initial synthesis are post-translational.} One of the most common modifications is phosphorylation where a phosphate group is attached to a specific amino acid in the protein. Phosphate groups contain three negative charges, so their addition can dramatically affect a protein's structural energetics and induce a conformational change. Thus, phosphorylation often plays a key role in toggling proteins between inactive and active states. In general, though, PTMs encompass a variety of chemical modifications with similarly diverse impacts on a protein's behavior. Because IDRs are generally exposed to their environment, they are frequent targets of PTMs. However, PTMs do not occur haphazardly in proteins but are instead targeted to binding sites created by sequential patterns of residues called short linear motifs (SLiMs). SLiMs are a general mechanism for mediating interactions with proteins, so while many SLiMs are targets of PTMs, others simply recruit binding partners. In contrast to the highly structured interfaces that characterize interactions with folded domains, however, SLiMs are usually no more than ten residues~\cite{Tompa2014}. Thus, their interactions are relatively weak and highly transient. As their flexibility makes SLiMs easily accessible, they are also enriched in IDRs, which in turn allows them to bind to many partners, sometimes simultaneously. IDRs therefore often act as hubs in complex regulatory networks by propagating signals from diverse sources or organizing binding partners into higher-order structures~\cite{Dunker2005, Wright2014}. Furthermore, these signals and interactions are easily tuned because PTMs can modulate their meanings and strengths, respectively, by modifying only a few residues. Thus, IDRs are like ``molecular computers'' that integrate complex data and respond accordingly to execute different genetic ``programs.''

>>>>

Unsurprisingly, disordered regions are characterized by a relative depletion of hydrophobic resi\-dues~\cite{Uversky2000}. However, there is no simple formula which accurately predicts disorder in a protein using only the hydrophobicity values of its constituent residues. Other factors, such as the number and distribution of charged residues, also impact a sequence's predisposition for disorder and its resulting conformational class~\cite{Mao2010, vanderLee2014, Das2015}. For example, sequences with high number of either positively or negatively charged residues, called polyelectrolytes, tend to form stiff rods because the like charges repel each other. However, if a sequence contains a high number of positively and negatively charged residues in roughly equal proportion, the sequence is called a strong polyampholyte, and its conformational class depends on the distribution of those charged residues. If the two classes of residues are segregated into separate blocks of like charges, they attract and form hairpins. However, if they are evenly distributed, the attractive and repulsive forces balance on average, and the sequence generally assumes expanded coil-like conformations. High numbers of polar amino acids, which are hydrophilic but not charged, are associated with semi-compact globules. Though their side chains, the portions which give each amino acid its unique identity, are hydrophilic, their interactions with water are not sufficient to overcome the tendency of the hydrophobic backbone, composed the donor and receptor sites common to all amino acids, to self-associate. Thus, like folded domains, the structures of disordered proteins are dictated by their sequences. However, because IDRs do not make stable contacts between specific residues in their chains, multiple sequences can generally correspond to a single conformational class.

The structural diversity of IDRs is directly related to their functional plasticity, and as IDRs are not confined to one conformation, they can interact with and bind to many possible partners. Often these partners are other proteins, but they can also be other macromolecules like DNA or even small molecules and ions. As a result of this adaptability, IDRs are enriched in proteins involved in cell signaling. Because cells are highly compartmentalized, they have a dizzying array of mechanisms to relay messages.\footnote{The most fundamental compartment is the cell, which roughly separates inside from outside and life from non-life. However, the cells of more complex organisms called eukaryotes contain additional subcompartments called organelles. Compartmentalization is essential in living systems because it confines and concentrates biochemical reactions that would be harmful if they occurred at the wrong place or time. However, it also introduces many complications because information in the form of physical or chemical signals cannot travel freely.} Many are mediated by interactions between proteins, which in turn create a change in the state of the system that propagates the signal. Often these state changes are chemical alterations made by one protein to another, which are termed post-translational modifications (PTMs).\footnote{A protein's amino acid sequence is encoded in a cell's DNA, and the process of reading that information to create a protein (from an intermediate molecule called RNA) is translation. Therefore, any modifications to a protein after its initial synthesis are post-translational.} One of the most common modifications is phosphorylation where a phosphate group is attached to a specific amino acid in the protein. Phosphate groups contain three negative charges, so their addition can dramatically affect a protein's structural energetics and induce a conformational change. Thus, phosphorylation often plays a key role in toggling proteins between inactive and active states. In general, though, PTMs encompass a variety of chemical modifications with similarly diverse impacts on a protein's behavior. Because IDRs are generally exposed to their environment, they are frequent targets of PTMs. However, PTMs do not occur haphazardly in proteins but are instead targeted to binding sites created by sequential patterns of residues called short linear motifs (SLiMs). SLiMs are a general mechanism for mediating interactions with proteins, so while many SLiMs are targets of PTMs, others simply recruit binding partners. In contrast to the highly structured interfaces that characterize interactions with folded domains, however, SLiMs are usually no more than ten residues~\cite{Tompa2014}. Thus, their interactions are relatively weak and highly transient. As their flexibility makes SLiMs easily accessible, they are also enriched in IDRs, which in turn allows them to bind to many partners, sometimes simultaneously. IDRs therefore often act as hubs in complex regulatory networks by propagating signals from diverse sources or organizing binding partners into higher-order structures~\cite{Dunker2005, Buljan2012, Wright2014}. Furthermore, these signals and interactions are easily tuned because PTMs can modulate their meanings and strengths, respectively, by modifying only a few residues. Thus, IDRs are like ``molecular computers'' that integrate complex data and respond accordingly to execute different genetic ``programs.''

### 2023-04-26 chapter2/chapter2.tex
Added citations to chapter 2 introduction paragraph 1

\section{Introduction}
Intrinsically disordered regions (IDRs) are segments of proteins which lack stable three-dimensional structures and instead exist as ensembles of rapidly interconverting conformations. As a result of this structural hetereogeneity, IDRs can interact with diverse binding partners. Often these interactions have high specificity but moderate affinity, which permits the efficient propagation of signals by rapid binding and dissociation. Furthermore, as IDRs readily expose their polypeptide chains, they are enriched in motifs for post-translational modifications which allow environmental or physiological conditions to modulate their interactions. Accordingly, IDRs often act as the ``hubs'' of complex signaling networks by integrating signals from diverse pathways and coordinating interactions. However, as IDRs are pervasive in eukaryotic proteomes, with an estimated X\% of proteins containing a disordered region of at least 30 residues, they are involved in diverse processes and phenomena, including EXAMPLE, EXAMPLE, and the formation of biomolecular condensates.

>>>>

\section{Introduction}
Intrinsically disordered regions (IDRs) are segments of proteins which lack stable three-dimensional structures and instead exist as ensembles of rapidly interconverting conformations. As a result of this structural hetereogeneity, IDRs can interact with diverse binding partners. Often these interactions have high specificity but moderate affinity, which permits the efficient propagation of signals by rapid binding and dissociation~\cite{Zhou2012, Wright2014}. Furthermore, as IDRs readily expose their polypeptide chains, they are enriched in recognition motifs for post-translational modifications which allow environmental or physiological conditions to modulate their interactions. Accordingly, IDRs often act as the ``hubs'' of complex signaling networks by integrating signals from diverse pathways and coordinating interactions~\cite{Dunker2005, Buljan2012}. However, as IDRs are pervasive in eukaryotic proteomes, with estimates of the fractions of disordered residues in the human, mouse, and fruit fly proteomes ranging between 22 and 24\%, they are involved in diverse functions~\cite{Piovesan2020, Piovesan2022}. These including EXAMPLE, EXAMPLE, and the formation of biomolecular condensates.
